---
title: "AAVSO-Zapper Summulator"
author: "David J Jackson"
date: "2/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(plotly)

```
```{r Import Data,echo=FALSE}
rm(list=ls())
AOri <- fread("../db/Alf_Ori-corrected.csv")
AOri$Ymd <- as.Date(paste(AOri$year, AOri$month, AOri$day, sep = "-"))
AOri <- AOri %>% select(Ymd,year,month,Vis)
summary(AOri)

```


## The Whole Enchilada: Jan - 1918 to Feb. 2020
```{r, echo=FALSE}
# AOri$MA <- forecast::ma(AOri$Vis,order=30)
plot_ly(AOri) %>% add_lines(x=~Ymd,y=~Vis) %>%
  layout(yaxis=list(autorange="reversed"))
```

## AAVSO Zapper Simulator (DataVerificaton)

Positive observations more than 1 magnitude brighter or fainter than the mean magnitude.

Make sure you are comparing observations made using the same color filter with each other (don't try to compare a CCDB observation with a visual one, for instance). 

Also,make sure you are comparing observations made at approximately the same time (for some stars, significant changes in behavior can occur in a very short time.)

  *  Fainter-than observations that are fainter than t5>he mean curve.
  *  Observations made during the "seasonal gap" of a star.
  *  For novae and supernovae, positive observations made before outburst.

```{r,echo=FALSE}
box_plot <-AOri %>% filter(Ymd >="2019-12-01") 
plot_ly(box_plot) %>% add_boxplot(x = ~Ymd, y = ~Vis) %>%
  layout(yaxis=list(autorange="reversed"))


```
```{r}
summary(box_plot)
```