---
title: "Alf (aka Betelgeuse) Ori Preditions(v1.3)"
subtitle: "Jan. 1964 - Mar.16 2020"
author: "Rodney Howe & DAvid  Jackson"
date: "2/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(lubridate)
library(plotly)
library(prophet)
library(forecast)
```

#  The American Association of Variable Star Observers (AAVSO) 

The AAVSO is an international non-profit organization of variable star observers whose mission is to enable anyone, anywhere, to participate in scientific discovery through variable star astronomy. We accomplish our mission by carrying out the following activities:

* observation and analysis of variable stars
* ollecting and archiving observations for worldwide access
* forging strong collaborations between amateur and professional astronomers
* promoting scientific research, education, and public outreach using variable star data.


## Analysis Visual Observations of Betelgeuse
```{r Import Alf Ori data, echo=FALSE}
rm(list=ls())
AOri <-fread("../db/Alf_Ori-corrected.csv")
AOri$Ymd <- as.Date(paste(AOri$year, AOri$month, AOri$day, sep = "-"))
AOri <- AOri %>% select(Ymd,year,month,day,Vis) 
AOri$Cts <- ifelse(AOri$Vis==0.00,0,1)
# Ori$Loess <-lowess(AOri$Vis,f=0.1)
summary(AOri)
```
```{r Calculate Daily Means}
 AOri1 <- AOri %>% select(Ymd,year,Vis) %>%
   group_by(Ymd) %>% summarise(Mean = mean(Vis),
                               Sum = sum(Vis),
                               Obs = n())
                                        

```

## First some basic Summary Stats
```{r echo=FALSE}
summary(AOri1)
```
```{r echo=FALSE}
thirty_days <- AOri %>% filter(Ymd >= "2020-01-18")
 ggplot(thirty_days,aes(Vis)) + geom_histogram() + labs(title="Histogram of Observed Magnatudes",subtitle="Jan. 18,2020 - Mar. 16,2020")
 
 ggplot(thirty_days) + geom_smooth(aes(x=Ymd,y=Vis)) +  
   scale_y_reverse() + geom_point(aes(x=Ymd,y=Vis)) + labs(title="Visual Obersvations for Jan 1,2020 - Mar. 16, 2020")
   
```
```{r echo=FALSE}
thirty_days$Ymd <- as.factor(thirty_days$Ymd)
ggplot(thirty_days) + geom_boxplot(aes(x=Ymd,y=Vis)) + coord_flip() +
  labs(title="Daily Visual Magnatudes", x="Date of Observation",y="Visual Magnatude",subtitle="Jan. 8 ,2020 -Feb. 16,2020 ") 
  # theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

##.... Seems to be a Spike in the Number of Observations
```{r echo=FALSE}
AOri1 %>% ggplot() + geom_col(aes(x=Ymd,y=Obs)) + 
      geom_smooth(aes(x=Ymd,y=Obs)) +
      labs(title="Betelgeuse: Number of Observation",
           subtitle="Jan. 1,1964 - Feb. 16, 2020",
           y="Number of Observation")
```


## Plot of (mean) Daily with Loess 
```{r echo=FALSE}
Fit <- as.data.frame(lowess(AOri1$Mean,f=0.1))
AOri2 <-cbind(AOri1,Fit$y)
colnames(AOri2) <-c("Ymd","Mean","Sum","Obs","Loess")
```
```{r echo=FALSE}

ggplot(data=AOri2,aes(x=Ymd,y=Mean,col="Visual")) +
  geom_line() +geom_line(aes(x=Ymd,y=Loess,col="Loess")) + 
  ggtitle("Alf Ori: Jan. 1, 1964 - Feb. 16,2020") +
  scale_y_reverse()
```
```{r echo=FALSE}
AOri2 %>% filter(Ymd >="2007-01-01") %>%
ggplot(aes(x=Ymd,y=Mean,col="Visual")) +
  geom_line() +geom_line(aes(x=Ymd,y=Loess,col="Loess")) + 
  ggtitle("Alf Ori: Jan. 1, 2009 - Feb. 16,2020") +
  scale_y_reverse()
```

## Calculate and Plot 60 Day Moving Avreage
```{r echo=FALSE}
 AOri1$MA <- ma(AOri1$Mean,order=60)
 ggplot(data=AOri1,aes(x=Ymd,y=MA,col="Vis")) +
  geom_line()  + 
   labs(title="Alf Ori (aka Betelgeuse) : Jan 1,1964 - Feb. 16,2020",subtitle="60 DAy Moving Average",y="Visual Magnitude",
                      ylab="Visual Magnitute") + 
   geom_hline(yintercept = 0.68) +
    scale_y_reverse()
```

## Use Facebook Prophet to Predict Magnitutde
```{r Predict Daily Alf Ori, echo=FALSE}
df <- AOri1 %>% select(Ymd,Mean)
colnames(df) <- c("ds","y")
m <- prophet(seasonality.mode="multiplicative")
m <- add_seasonality(m, name="half_year_cycle", period=180.25 * 1,fourier.order=5)
m <- fit.prophet(m, df)
future <- make_future_dataframe(m,periods=500,freq="day",include_history = TRUE)
forecast <- predict(m, future)

```
```{r echo=FALSE}
plot(m, forecast) +ggtitle("Vis Daily: Jan. 1964 - Mar. 2020") +
 ylab("Predicted Days w/ Vis") +   xlab("Years" ) +  scale_y_reverse()
```
```{r echo=FALSE}
fcast <- forecast %>% select(ds,yhat,yhat_lower,yhat_upper) %>%
  filter(ds >="2019-03-06")
fcast$ds <- as.Date(fcast$ds)
Actual <- AOri1 %>%  filter(Ymd >="2007-01-01" & Ymd <="2019-10-01")
ggplot(fcast) + geom_line(aes(x=ds,y=yhat,col="Prediction")) +
  geom_line(data=Actual,aes(x=Ymd,y=MA,col="Moving Average")) + 
#  geom_line(data=fcast,aes(x=ds,y=yhat_upper)) +
#  geom_line(data=fcast,aes(x=ds,y=yhat_lower)) +
  scale_y_reverse() +
  labs(title = "Alf (aka Betelgeuse) Ori: Actual vs Predicted ", 
       subtitle = "(Actual = 60 Day Moving Average)",y="Visual Magnitude",x="Year")
```
