---
title: "Rodney's Solar ISN Data"
author: "David Jackson"
date: "3/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(RSQLite)
db <- dbConnect(SQLite(), dbname="./RHOWE.sqlite3")
```

## Rodney Howe's  Data

If you have the time could you make an .Rmd file that loads these two data sets into a sqlite3 database?  Just the .csv files?  

http://www.observations.be/silso/groupnumberv3

Archive of raw group counts
This is the current master archive of all base group observations used as input data for building the current group number series. 
Reference: Revised collection of the number of sunspot groups from 1610 to 2010.J.M. Vaquero, L. Svalgaard, V.M.S. Carrasco, F. Clette, L. Lefevre, M.C. Gallego, R. Arlt, A.J.P. Aparicio, J.-G. Richard, and R. Howe, 2016: "Revised collection of the number of sunspot groups from 1610 to 2010", Solar Physics, submitted

### Read in Changes

```{r}
changes <- read.table("./DATA/GNchanges.txt", sep = '\t',header = TRUE)
```


### Read Observations

```{r}
observations <- read.csv("./DATA/GNobservations.csv")
observations$Ymd <- as.Date(paste(observations$YEAR, observations$MONTH, observations$DAY, sep = "-"))
```

### Read Observers

```{r}
observers <- read.csv("./DATA/GNobservers.csv")
```

### Create and Update SQLite DB 

```{r}
observations$Ymd <- as.character(observations$Ymd)
dbWriteTable(db, "observations",observations ,overwrite=TRUE)
```
```{r}
dbWriteTable(db, "observers",observers ,overwrite=TRUE)
```

